Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-29 manuscript/manuscript.tex
Edit the conclusions
Emphasize that the method is general purpose. We sacrifice speed for
usability and accuracy.
Added some more speculation about using the density-based discretization
with other methods.
Talk a bit about the Neuquen application and how the errors we see there
would translate into bad gravity inversions.

I removed the Gzz part because it felt very out of place. We don't talk
about it anywhere and have nothing to show. The only people who would
ask are the reviewers so we should include that in our response instead.

Both the \citet{Lin2018} and the \citet{Fukushima2018} studies limit the radial density
variation to polynomial functions.
While most continuous and smooth functions can be approximated by piecewise linear
functions, the choice of a discretization interval is neither straight forward nor
automatic.
Furthermore, it is well known that the use of high-degree polynomials to approximate a
highly variable function produces unstable results when extrapolating beyond the data
domain.
These shortcomings could make piecewise linear or high-degree polynomial density
functions unwieldy for non-linear gravity inversions if the density function in depth
is highly variable.

@@

We have developed a new methodology to compute the gravitational fields
generated by a tesseroid with a density given by a continuous function of depth.
It numerically solves the integrals that define the gravitational potential and
its gradient through the Gauss-Legendre Quadrature (GLQ).
The accuracy of the numerical integration is automatically controlled by a two
dimensional adaptive discretization algorithm and a new density-based discretization
algorithm.
The former divides the tesseroid in half if the ratio of the distance to the computation
point and the latitudinal or the longitudinal size of the tesseroid is lower than
a predefined distance-size ratio $D$.
This algorithm minimises the integration error when the computation point is close to
the tesseroid.
Nevertheless, the adaptive discretization alone is not sufficient to guarantee the
accuracy of the method in case of tesseroids with variable density.

To overcome this challenge, we have developed a density-based discretization algorithm
that divides the tesseroid on the points at which the maximum variations of the density
function take place.
The density-based discretization is performed before adaptive discretization as a type
of pre-processing step.
The number of divisions performed, and thus the accuracy of the computation, is
controlled by the delta ratio $\delta$.
This new algorithm is intended to minimise the error due to the inability of
the GLQ to produce precise approximations of density functions with sharp variations.

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimal values for these
parameters.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimal values were a linear, an exponential
and a sinusoidal function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analysed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
And the error for sinusoidal functions with different wavelengths.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1 and 2.5 for the potential an its gradient, respectively.
While values of $\delta$ equal to 0.1 and 0.01 are needed in order to guarantee a 0.1\%
accuracy for an exponential and a sinusoidal density, respectively.
These results could be extrapolated: a $\delta$ value equal to 0.01 is needed in order
to achieve the accuracy threshold in case of an arbitrary continuous density function.
Although we encourage to perform similar accuracy tests for functions other than the
ones we studied, this results could be safely generalized to other common depth-density
relations among geophysical applications, like quadratic functions, without loss of
accuracy.

% Moved here from the discussion. The conclusions are the place to talk about future
% work and speculations.
We have also carried out $D$ and $\delta$ determinations for the vertical component of
the Marussi tensor ($g_{zz}$) for linear and exponential densities.
We didn't include these results because our methodology wouldn't achieve the 0.1\%
accuracy threshold for the general case of such density functions.
Moreover, the values of $D$ and $\delta$ needed for the cases at which the accuracy was
below 0.1\% implied a high number of discretizations, thus greatly increasing the
computation time.
In summary, we found the computation of the Marussi tensor components in case of
variable density tesseroids a subject for future research.

>>>>

Both the \citet{Lin2018} and the \citet{Fukushima2018} studies limit the radial density
variation to polynomial functions.
While most continuous and smooth functions can be approximated by piecewise linear
functions, the choice of a discretization interval is neither straight forward nor
automatic.
Furthermore, it is well known that the use of high-degree polynomials to approximate a
highly variable function produces unstable results when extrapolating beyond the data
domain.
These shortcomings could make piecewise linear or high-degree polynomial density
functions unwieldy for non-linear gravity inversions \citep[e.g.][]{Uieda2017} if the
density function in depth is highly variable.

@@

We have developed a new methodology to compute the gravitational potential and its
gradient for a tesseroid with a density given by a continuous function of radius.
It numerically solves the volume integrals through Gauss-Legendre Quadrature (GLQ) by
including the density function in the numerical integration.
By implementing the algorithm in the dynamic Python programming language, users can
define their own density function to supply to the software.
This allows the use of any arbitrary continuous function without modification to the
method or software.
The accuracy of the numerical integration is automatically controlled by a two
dimensional adaptive discretization algorithm and a new density-based discretization
algorithm.
The former divides the tesseroid in half if the ratios of the distance to the
computation point and the latitudinal or the longitudinal size of the tesseroid are
smaller than a predefined distance-size ratio $D$.
This algorithm minimises the integration error when the computation point is close to
the tesseroid.
Nevertheless, the adaptive discretization alone is not sufficient to guarantee the
accuracy of the method in case of tesseroids with non-linear density functions.
To overcome this challenge, the density-based discretization algorithm
divides the tesseroid along the radial dimension on the points at which the maximum
variations of the density function take place.
The number of radial divisions performed, and thus the accuracy of the computation, is
controlled by the $\delta$ parameter.
This new algorithm is intended to minimise the error due to the inability of
the GLQ to produce precise approximations of density functions with sharp variations.

We empirically determined optimal values for $D$ and $\delta$ using common model and
computation grid configurations and density functions.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimal values were a linear, an
exponential, and a sinusoidal function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
Through this analysis, we have obtained optimal values for the distance-size ratio $D$
of 1 and 2.5 for the potential an its gradient, respectively.
We analysed the error for exponential functions ranging from smooth to sharp and
sinusoidal function of different wavelengths to test the accuracy of the density-based
discretization.
Values of $\delta$ equal to 0.1 are sufficient to guarantee a 0.1\% accuracy for most
realist cases tested.
These results could be extrapolated to other continuous density functions that are
sufficiently smooth, as is the case for most common geophysical applications.
Smaller values of $\delta$ can used for more highly variable density functions to
increase the accuracy of the integration accordingly.

Some computational performance is sacrificed in order to obtain an accurate and general
purpose method.
The density-based discretization adds to the computation time by increasing
the number of GLQ integrations.
Likewise, allowing users to supply custom density functions incurs an overhead cost
because of the added function evaluations.
It is also impossible to optimize the source code and formulas without knowledge of the
specific density function.
Nonetheless, the density-based discretization is independent of the adaptive
discretization and GLQ integration.
It can be viewed as a type of pre-processing step that can be combined with other more
specialized integration methods.

An application to modeling the Neuqu\'en Basin, Argentina, demonstrates that the effects
of sediment compaction cannot be ignored.
When compared to an exponential density function, a homogeneous density with depth would
result in an error of up to -8 mGal, while a linear density function would result in up
to -6 mGal error.
Accurate and robust forward modeling is a key component to any gravity inversion method.
Errors of this magnitude could result in significant overestimation of sedimentary
basin thickness, for example.
